---
title: 'Introduction to R: mtcars'
author: "Sean Cho"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load dataset

We will be working with the `mtcars` dataset. We will first load the dataset and then find out more about the dataset.

We can use `?` or `help` to get information about `function`s. We can also use them to find out more about datasets, including `mtcars`.

```{r mtcars}
data('mtcars')
?mtcars

## A data frame with 32 observations on 11 (numeric) variables.
##
## [, 1]	mpg	Miles/(US) gallon
## [, 2]	cyl	Number of cylinders
## [, 3]	disp	Displacement (cu.in.)
## [, 4]	hp	Gross horsepower
## [, 5]	drat	Rear axle ratio
## [, 6]	wt	Weight (1000 lbs)
## [, 7]	qsec	1/4 mile time
## [, 8]	vs	Engine (0 = V-shaped, 1 = straight)
## [, 9]	am	Transmission (0 = automatic, 1 = manual)
## [,10]	gear	Number of forward gears
## [,11]	carb	Number of carburetors
```

## Examine dataset

We can look at the structure of `mtcars` using the `str()`.

```{r mtcars_examine}
str(mtcars)
```

We can see that the `data.frame` has 11 variables, or columns, and 32 observations, or rows. Although all of these are numeric, we know that `vs` and `am` are binary columns of whether the car has a V-shaped engine and whether the car has an automatic or manual transmission.

Next, we will use `summary` to summarise the `mtcars` data frame.

```{r mtcars_summary}
summary(mtcars)
```

We see that `mtcars$mpg` ranges from 10.40 to 33.90 with a mean of 20.09 and a median of 19.20. 

Now, we'll take a look at the first few rows of mtcars.

```{r mtcars_head}
head(mtcars)
```

## Exploratory analysis

By using the `plot` function on a `data.frame`, we can make pair-wise scatterplots for the columns in the `data.frame`. Here, we will `plot` the first five columns of `mtcars`.

```{r mtcars_plot}
plot(mtcars[,1:5])
```

We can see a few relationships between `mpg` and the other variables. One of them is `hp`, or horsepower. 

Does this make sense? Do we expect that miles/gallon is inversely related to horsepower?
Yes. We would expect that a car with greater horsepower would tend to be less efficient.

## Evaluating categorical variables

There are several categorical variables in the `mtcars` dataset, including `cyl`, `gear`, and `carb`. We can examine the relationship between `mpg` and `cyl` and visualise that using a boxplot.

```{r mtcars_boxplot}
boxplot(mtcars$mpg ~ mtcars$cyl, xlab = 'Number of Cylinders', ylab = 'MPG')
```

From the boxplot, we can observe that there is a stepwise decrease in mpg with increasing cylinders. There is very little overlap between the boxplots across cylinders and we can test that if there are statistically significant differences.

We will run two tests. (1) an ANOVA to identify if any of the `cyl` groups have different mpg values and (2) a pairwise t-test to identify differences across groups.

```{r mtcars_anova}
## ANOVA analysis
anova(aov(mpg ~ cyl, data = mtcars))

## pairwise t-test
pairwise.t.test(mtcars$mpg, g = mtcars$cyl)
```

In this sample, we can conclude that there are pairwise differences of mpg across number of cylinders of the car's engine.

## Comparing numerical variables

We will continue to explore the relationship between mpg and horsepower that we observed earlier by making a scatterplot of the two variables.

```{r mtcars_hp}
## plot( y ~ x , data = dataset)
plot(mpg ~ hp, data = mtcars, ylab = 'MPG', xlab = 'Horsepower')

## plot(x = mtcars$mpg, y = mtcars$mpg) will work as well
```

We can observe what appears to be an inverse relationship between mpg and horsepower. We can fit a linear model that 

```{r mtcars_linear}
## lm( y ~ x1 + x2 + ... + xn , data = dataset)
summary(lm(mpg ~ hp, data = mtcars))
```

From this model, we observe that the estimate is that for every 1 point increase of horsepower, there is a decrease of -0.06823 mpg.

So now we know that there is an inverse relationship between mpg and horsepower on top of the relationship between mpg and the number of cylinders.
Let's include all that information in our plot.

```{r mtcars_abline}
## create colors
mapcolors <- c('4'='black','6'='skyblue3','8'='orange')
mtcolors <- mapcolors[as.character(mtcars$cyl)]

## model
mtmodel <- lm(mpg ~ hp, data = mtcars)

## make basic plot
plot(mpg ~ hp, data=mtcars, col = mtcolors, pch = 16, xlab = 'Horsepower', ylab = 'MPG',
     main = 'MPG ~ Horsepower')
## add fitted line
abline(mtmodel)
## add legend
legend('topright', legend = c(4,6,8), col = c('black','skyblue3','orange'), pch = 16, title = 'Cylinder')
```

## Multiple regression

We will explore the relationship between `vs`, V-shaped engine, and mpg. Let's first fit a model to assess that.

```{r mtcars_multiple_first}
unique(mtcars$vs)
## V-shaped engine
summary(lm(mpg ~ vs ,data = mtcars))
```

From this analysis, we can see that there is a significant difference between the mpg of cars with a V-shaped versus an inline engine. Do we believe that this relationship is real?
Maybe. Or maybe there are other covariates that we have not considered.

In this dataset, we will show here that cars with V-shaped engines differ from those with inline engines in horsepower and weight, and that difference captures more mpg variability than engine shape.

```{r mtcars_multiple}
boxplot(mtcars$hp ~ mtcars$vs)
boxplot(mtcars$wt ~ mtcars$vs)
## V-shaped engine and hp and weight
summary(lm(mpg ~ vs + hp + wt ,data = mtcars))
```
